Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-04-11 brms_meeg.qmd
new sections in the intro

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
abstract: "Time-resolved electrophysiological measurements such as those obtained through magneto- or electro-encephalography (M/EEG) offer a unique window into the neural activity underlying cognitive processes. Researchers are often interested in determining whether and when these signals differ across experimental conditions or participant groups. The conventional approach involves mass-univariate statistical testing across time and space, followed by corrections for multiple comparisons such as cluster-based inference. While effective for controlling error rates at the cluster-level, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual time points to clusters, it makes difficult to draw precise conclusions about the onset or offset of observed effects. Here, we introduce a *model-based* approach for analysing M/EEG timeseries such as event-related potentials (ERPs) or decoding performance over time. Our approach leverages Bayesian generalised additive multilevel models, providing posterior probabilities that an effect is above zero (or above chance) at each time point, while naturally accounting for temporal dependencies and between-subject variability. Using both simulated and actual M/EEG datasets, we demonstrate that this approach substantially outperforms conventional methods in estimating the onset and offset of neural effects, yielding more precise and reliable results. We provide an R package implementing the method and describe how it can be integrated into M/EEG analysis pipelines using MNE-Python."
keywords: [EEG, MEG, cluster-based inference, simulation, multiple comparisons, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: man # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
          \usepackage{booktabs, caption, longtable, colortbl, array}
---

@@

The conventional approach involves mass-univariate statistical testing through time and/or space followed by some form or correction for multiple comparisons with the goal of maintaining the familywise error rate (FWER) or false discovery rate (FDR) at the nominal level (e.g., 5%). Cluster-based inference is the most common way of achieving this sort of error control in the M/EEG literature, being the recommended approach in several software programs [e.g., `EEGlab` or `MNE-Python`, @delorme2004; @gramfort2013]. While effective for controlling error rates, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual time points to clusters, it prevents the ability to draw precise conclusions about the onset or offset of observed effects [@maris2007; @sassenhagen2019]. As pointed out by @maris2007; "there is a conflict between this interest in localized effects and our choice for a global null hypothesis: by controlling the FA [false alarm] rate under this global null hypothesis, one cannot quantify the uncertainty in the spatiotemporal localization of the effect". Even worse, @rosenblatt2018 note that cluster-based inference suffers from low spatial resolution: "Since discovering a cluster means that 'there exists at least one voxel with an evoked response in the cluster', and not that 'all the voxels in the cluster have an evoked response', it follows that the larger the detected cluster, the less information we have on the location of the activation." As a consequence, cluster-based inference is expected to perform poorly for localising the onset of M/EEG effects; a property that was later demonstrated in simulations studies [e.g., @rousselet_using_2025; @sassenhagen2019].

To overcome the limitations of cluster-based inference, we introduce a novel *model-based* approach for precisely localising M/EEG effects in time, space, and other dimensions. The proposed approach, based on Bayesian generalised additive multilevel models, allows quantifying the posterior probability of effects being above chance at the level of timesteps, sensors, voxels, or other dimensions, while naturally taking into account spatiotemporal dependencies present in M/EEG timeseries. We compare the performance of the proposed approach to well-established alternative methods using both simulated and actual M/EEG data and show that it significantly outperforms alternative methods in estimating the onset and offset of M/EEG effects.

## Cluster-based inference

The issue with multiple comparisons... Different methods exist to control the family-wiser error rate (FWER), defined as the type-1 error rate (false positive) over an ensemble (family) of tests... for instance the Bonferroni correction [@dunn1961], however this method is generally overconservative as it assumes statistical independence of tests, an assumptions that is clearly violated in the context of M/EEG timeseries with spatiotemporal dependencies... or the false discovery rate (FDR), defined as the proportion of false positive among positive tests [e.g., @benjamini1995; @benjamini2001]...

Some useful references to be discussed [@ehinger_unfold_2019; @frossard2021; @frossard2022]...

On popular technique to account for spatiotemporal dependencies while controlling the FWER is cluster-based inference. Description of cluster-based approaches [see @sassenhagen2019]... A typical cluster-based inference consists of two successive steps. First, clusters are defined as sets of contiguous voxels, channels, and/or timesteps, whose intensity/activity exceeds some predefined threshold. Clusters are then characterised by their height (i.e., maximal value), extent (number of constituent elements), or some combination of both, such at its "mass", for instance by summing the statistics within a cluster, an approach refereed to as "cluster mass" [@maris2007; @pernet2015]. Then, the null hypothesis is tested by assessing the probability.. As different cluster-forming thresholds lead to clusters with different spatial or temporal extent, this initial threshold modulates the sensitivity of the subsequent permutation test. The threshold-free cluster enhancement method (TFCE) was introduced by @smith2009 to overcome this choice of an arbitrary threshold.

In brief, the TFCE method works as follows. Instead of picking an arbitrary cluster-forming threshold (e.g., $t=2$), we try all (or many) possible thresholds in a given range and check whether a given timestep/voxel belongs to a significant cluster under any of the set of thresholds. Then, instead of using cluster mass (e.g., the sum of squared t-values within the cluster), we use a weighted average between the cluster extend ($e$, how broad is the cluster, that is, how many connected samples it contains) and the cluster height ($h$, how high is the cluster, that is, how large is the test statistic). The TFCE score at each timestep/voxel $t$ is given by:

$$
\text{TFCE(t)} = \int_{h=h_{0}}^{h_{t}} e(h)^{E} h^{H} \mathrm{d}h
$$

where $h_{0}$ is typically $0$ and parameters $E$ and $H$ are set a priori (typically to 0.5 and 2, respectively) and control the influence of the extend and height on the TFCE. Note that in practise, this intergral is approximated by a sum over small $h$ increments. Then, p-value for timestep/voxel $t$ is computed by comparing it TFCE with the null distribution of TFCE values. For each permuted signal, we keep the maximal value over the whole signal for the null distribution of the TFCE. The TFCE combined with permutation (assuming a large enough number of permutations) has been shown to provide accurate type 1 FWER [e.g., @pernet2015]. However, previous simulation work showed that cluster-based methods (including the TFCE method) perform poorly in localising the onset of M/EEG effects [e.g., @rousselet_using_2025; @sassenhagen2019].

To sum up this section, cluster-based inference main limitations is that it provides inference at the cluster level only, not allowing inference at the level of timesteps, sensors, etc. As a consequence, it does not allow inferring the localisation of effects [for more details on cluster-based inference, see for instance @maris2011; @maris2007; @sassenhagen2019]. In the following, we briefly review previous M/EEG modelling work. Then, we provide a brief introduction to generalised additive models (GAMs) and Bayesian generalised additive multilevel models (BGAMMs) to illustrate how these models can be used to precisely localise the onset and offset of M/EEG effects.

@@

From @dimigen2021: Recently, spline regression has been applied to ERPs [e.g., @hendrix2017; @kryuchkova2012]... GAMMs for EEG data [@abugaber2023; @meulman2015; @meulman2023]...

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. To clarify the terminology at this stage: *splines* are functions composed of simpler functions. These simpler functions are basis functions (e.g., cubic polynomial, thin-plate) and the set of basis functions is a *basis*. Each basis function gets its coefficient and the resultant spline is the sum of these weighted basis functions (@fig-intro-gam). Splines coefficients are penalised (usually through the squared of the smooth functions' second derivative) in a way that can be interpreted, in Bayesian terms, as a prior on the "wiggliness" of the function. In other words, more complex (wiggly) basis function are penalised.

@@

However, the previous model only included constant (fixed) effects, thus not properly accounting for between-participant variability. We next fit a multilevel version of the BGAM [BGAMM, see @nalborczyk2019]. Although it is possible to fit a BGAMM at the single-trial level, we present a computationally-lighter version of the model that is fitted directly on by-participant summary statistics (mean and SD), similar to what is done in meta-analysis.

@@

# fitting the BGAMM
meta_gam <- brm(
    # using by-participant SD of ERPs across trials
    eeg_mean | se(eeg_sd) ~
        condition + s(time, bs = "cr", k = 20, by = condition) +
        (1 | participant),
    data = summary_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/meta_gam.rds"
    )
```

@@

We then compute the posterior probability of the slope for `condition` being above $0$ (@fig-post-prob-ratio, left). From this quantity, we then compute the ratio of posterior probabilities (i.e., $p/(1-p)$) and visualise the timecourse of this ratio superimposed with the conventional thresholds on evidence ratios (@fig-post-prob-ratio, right). Note that a ratio of 10 means that the probability of the difference being above 0 is 10 times higher than the probability of the difference not being above 0, given the data, the priors, and other model's assumptions. Thresholding the posterior probability ratio thus provides a model-based approach for estimating the onset and offset of M/EEG effects.

@@

@@

@@

We then compared the ability of the BGAMM to correctly estimate the onset and offset of the ERP difference to other widely-used methods. First, we conducted mass-univariate t-tests (thus treating each timestep independently) and identified the onset and offset of the ERP difference as the first and last values crossing an arbitrary significance threshold ($\alpha = 0.05$). We then followed the same approach but after applying different forms of multiplicity correction to the p-values. We compared two methods that control the false discovery rate (FDR) [i.e., BH95, @benjamini1995; and BY01, @benjamini2001], one method that controls the family-wise error rate (FWER) [i.e., Holm–Bonferroni method, @holm1979], and two cluster-based permutation methods [permutation with a single cluster-forming threshold and threshold-free cluster enhancement, TFCE, @smith2009]. The BH95, BY01, and Holm corrections were applied to the p-values using the `p.adjust()` function in `R`. The cluster-based inference was implemented using a cluster-sum statistic of squared t-values, as implemented in `MNE-Python` [@gramfort2013], called via `r pkrt("reticulate")`. We also compared these estimates to the onset and offset as estimated using the binary segmentation algorithm, as implemented in `r pkrt("changepoint")`, and applied directly to the squared t-values [as in @rousselet_using_2025]. @fig-corrections illustrates the onsets and offsets estimated by each method on a single simulated dataset and shows that all methods systematically overestimate the true onset and underestimate the true offset.

@@

To assess the accuracy of group-level onset estimation, the various methods were compared using the bias (i.e., median(estimated-true)), median absolute error (MAE), root mean square error (RMSE), variance, and median absolute deviation (MAD) of onset/offset estimates computed on 1,000 simulated datasets. As in @rousselet_using_2025, each participant was assigned a random onset between 150 and 170ms.

@@

To complement the simulation study, we evaluated the performance of the various methods on actual MEG data [decoding results from @nalborczyk:inprep]. In this study, we conducted time-resolved multivariate pattern analysis (MVPA, also known as decoding) of MEG data during reading tasks. As a result, we obtain a timecourse of decoding performance (ROC AUC), bounded between 0 and 1, for each participant (for a total of 32 participants). Next, we wanted to *test* whether the group-level average decoding accuracy is above chance (i.e., 0.5) at each timestep (@fig-decoding-data). To achieve this, we fitted a BGAM as introduced previously, but we replaced the $\mathrm{Normal}$ likelihood function by a $\mathrm{Beta}$ one to account for the bounded nature of AUC values (between 0 and 1) [for a tutorial on Beta regression, see @coretta2025].

Note that although we chose a basis dimension of $k=50$, which seems appropriate for the present data, this choice should be adapted according to the properties of the modelled data (e.g., signal-to-noise ratio, prior low-pass filtering, sampling rate, etc) and should be assessed by the usual model checking tools (e.g., posterior predictive checks, see also @sec-basis). To better distinguish signal from noise, we also defined a region of practical equivalence [ROPE, @kruschke2017], defined as the chance level plus the standard deviation of the (group-level average) decoding performance during the baseline period.

@@

```{r tbl-simulation-results, eval = TRUE, echo = FALSE, results = "asis"}
#| tbl-cap: "Summary statistics of the onset and offset estimates for each method (ordered by MAE)."

@@

>>>>

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
abstract: "Time-resolved electrophysiological measurements such as those obtained through magneto- or electro-encephalography (M/EEG) offer a unique window into the neural activity underlying cognitive processes. Researchers are often interested in determining whether and when these signals differ across experimental conditions or participant groups. The conventional approach involves mass-univariate statistical testing across time and space, followed by corrections for multiple comparisons such as cluster-based inference. While effective for controlling error rates at the cluster-level, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual time points to clusters, it makes difficult to draw precise conclusions about the onset or offset of observed effects. Here, we introduce a *model-based* approach for analysing M/EEG timeseries such as event-related potentials (ERPs) or decoding performance over time. Our approach leverages Bayesian generalised additive multilevel models, providing posterior probabilities that an effect is above zero (or above chance) at each time point, while naturally accounting for temporal dependencies and between-subject variability. Using both simulated and actual M/EEG datasets, we demonstrate that this approach substantially outperforms conventional methods in estimating the onset and offset of neural effects, yielding more precise and reliable results. We provide an R package implementing the method and describe how it can be integrated into M/EEG analysis pipelines using MNE-Python."
keywords: [EEG, MEG, cluster-based inference, simulation, multiple comparisons, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# see available font via systemfonts::system_fonts()
# monofont: Monaco
# monofont: Courier
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
suppress-title-introduction: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
          \usepackage{booktabs, caption, longtable, colortbl, array}
          \usepackage{setspace}
          \onehalfspacing
---

@@

The conventional approach involves mass-univariate statistical testing through time and/or space followed by some form or correction for multiple comparisons with the goal of maintaining the familywise error rate (FWER) or the false discovery rate (FDR) at the nominal level (e.g., 5%). Cluster-based inference is the most common way of achieving this sort of error control in the M/EEG literature, being the recommended approach in several software programs [e.g., `EEGlab`, @delorme2004; `MNE-Python`,  @gramfort2013]. While effective for controlling error rates, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual datapoints (e.g., timesteps, sensors, voxels) to clusters, it prevents the ability to draw precise conclusions about the spatiotemporal localisation of such effects [@maris2007; @sassenhagen2019]. As pointed out by @maris2007: "there is a conflict between this interest in localized effects and our choice for a global null hypothesis: by controlling the FA [false alarm] rate under this global null hypothesis, one cannot quantify the uncertainty in the spatiotemporal localization of the effect". Even worse, @rosenblatt2018 note that cluster-based inference suffers from low spatial resolution: "Since discovering a cluster means that 'there exists at least one voxel with an evoked response in the cluster', and not that 'all the voxels in the cluster have an evoked response', it follows that the larger the detected cluster, the less information we have on the location of the activation." As a consequence, cluster-based inference is expected to perform poorly for localising the onset of M/EEG effects; a property that was later demonstrated in simulations studies [e.g., @rousselet_using_2025; @sassenhagen2019].

To overcome the limitations of cluster-based inference, we introduce a novel *model-based* approach for precisely localising M/EEG effects in time, space, and other dimensions. The proposed approach, based on Bayesian generalised additive multilevel models, allows quantifying the posterior probability of effects being above chance at the level of timesteps, sensors, voxels, etc, while naturally taking into account spatiotemporal dependencies present in M/EEG data. We compare the performance of the proposed approach to well-established alternative methods using both simulated and actual M/EEG data and show that it significantly outperforms alternative methods in estimating the onset and offset of M/EEG effects.

## Statistical errors and cluster-based inference

The issues with multiple comparisons represent a common and well-recognised danger in neuroimaging and M/EEG research where the collected data allows for a multitude of potential hypothesis tests and is characterised by complex structures of spatiotemporal dependencies. The probability of obtaining at least one false positive in an ensemble (family) of $m$ tests (i.e., the FWER) is computed as $1-\left(1-\alpha\right)^{m}$ (for $m=10$ tests and $\alpha=0.05$, it is approximately equal to $0.4$). Different methods exist to control the FWER (i.e., bring it back to $\alpha$). Most methods apply a simple correction to series of "raw" $p$-values issued from univariate statistical tests (e.g., t-tests). For instance, the Bonferroni correction [@dunn1961] consists in setting the significance threshold to $\alpha/m$, or equivalently, multiplying the "raw" $p$-values by $m$ and using the standard $\alpha$ significance threshold. This method is generally overconservative (i.e., under-powered) as it assumes statistical independence of the tests, an assumption that is clearly violated in the context of M/EEG timeseries characterised by massive spatiotemporal dependencies. Some alternative methods aims at controlling the FDR, defined as the proportion of false positive *among positive tests* [e.g., @benjamini1995; @benjamini2001]. However, a major limitation of both types of corrections is that they do not take into account the spatial and temporal information contained in M/EEG data.

A popular technique to account for spatiotemporal dependencies while controlling the FWER is cluster-based inference [@bullmore1999; @maris2007]. A typical cluster-based inference consists of two successive steps [for more details on cluster-based inference, see for instance @frossard2022; @maris2011; @maris2007; @sassenhagen2019]. First, clusters are defined as sets of contiguous timesteps, sensors, voxels, etc, whose activity, summarised by some test statistic (e.g., a $t$-value), exceeds a predefined threshold (e.g., the 95th percentile of the parametric null distribution). Clusters are then characterised by their height (i.e., maximal value), extent (number of constituent elements), or some combination of both, for instance by summing the statistics within a cluster, an approach referred to as "cluster mass" [@maris2007; @pernet2015]. Then, the null hypothesis is tested by computing a $p$-value for each identified cluster by comparing its mass with the null distribution of cluster masses (obtained via permutation). As alluded previously, a significant cluster is a cluster which contains *at least one* significant time-point. As such, it would be incorrect to conclude, for instance, that the timestep of a significant cluster is the first moment at which some conditions differ [@frossard2022; @sassenhagen2019]. In other words, because the inference is performed at the second step (i.e., once clusters have been formed), it prevents any conclusion to be made about individual datapoints (e.g., timesteps, sensors, etc).

As different cluster-forming thresholds lead to clusters with different spatial or temporal extent, this initial threshold modulates the sensitivity of the subsequent permutation test. The threshold-free cluster enhancement (TFCE) method was introduced by @smith2009 to overcome this choice of an arbitrary threshold. In brief, the TFCE method works as follows. Instead of picking an arbitrary cluster-forming threshold (e.g., $t=2$), the methods consist in trying all (or many) possible thresholds in a given range and checking whether a given datapoint (e.g., timestep, sensor, voxel) belongs to a significant cluster under any of the set of thresholds. Then, instead of using cluster mass, one uses a weighted average between the cluster extend ($e$, how broad is the cluster, that is, how many connected samples it contains) and the cluster height ($h$, how high is the cluster, that is, how large is the test statistic). The TFCE score at each timestep $t$ is given by:

$$
\text{TFCE(t)} = \int_{h=h_{0}}^{h=h_{t}} e(h)^{E} h^{H} \mathrm{d}h
$$

where $h_{0}$ is typically $0$ and parameters $E$ and $H$ are set a priori (typically to 0.5 and 2, respectively) and control the influence of the extend and height on the TFCE. Note that in practise, this integral is approximated by a sum over small $h$ increments. Then, a p-value for each timestep $t$ is computed by comparing its TFCE with the null distribution of TFCE values (obtained via permutation). For each permuted signal, we keep the maximal value over the whole signal for the null distribution of the TFCE. The TFCE combined with permutation (assuming a large enough number of permutations) has been shown to provide accurate FWER control [e.g., @pernet2015]. However, further simulation work showed that cluster-based methods (including TFCE) perform poorly in localising the onset of M/EEG effects [e.g., @rousselet_using_2025; @sassenhagen2019].

To sum up, the main limitation of cluster-based inference is that it allows for inference at the cluster level only, not allowing inference at the level of timesteps, sensors, etc. As a consequence, it does not allow inferring the precise spatial and temporal localisation of effects. In the following, we briefly review previous M/EEG modelling work. Then, we provide a short introduction to generalised additive models (GAMs) and Bayesian generalised additive multilevel models (BGAMMs) to illustrate how these models can be used to precisely localise the onset and offset of M/EEG effects.

@@

From @dimigen2021: Recently, spline regression has been applied to ERPs [e.g., @hendrix2017; @kryuchkova2012]... GAMMs for EEG data [@abugaber2023; @meulman2015; @meulman2023]... See also the UNFOLD toolbox [@ehinger_unfold_2019]...

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. To clarify the terminology at this stage: *splines* are functions composed of simpler functions. These simpler functions are basis functions (e.g., cubic polynomial, thin-plate) and the set of basis functions is a *basis*. Each basis function gets its coefficient and the resultant spline is the sum of these weighted basis functions (@fig-intro-gam). Splines coefficients are penalised (usually through the squared of the smooth functions' second derivative) in a way that can be interpreted, in Bayesian terms, as a prior on the "wiggliness" of the function. In other words, more complex (wiggly) basis functions are penalised.

@@

However, the previous model only included constant (fixed) effects, thus not properly accounting for between-participant variability. We next fit a multilevel version of the BGAM [BGAMM, for an introduction to Bayesian multilevel models in `brms`, see @nalborczyk2019]. Although it is possible to fit a BGAMM at the single-trial level, we present a computationally-lighter version of the model that is fitted directly on by-participant summary statistics (mean and SD), similar to what is done in meta-analysis.

@@

# fitting the BGAMM
meta_gam <- brm(
    # using by-participant SD of ERPs across trials
    eeg_mean | se(eeg_sd) ~
        condition + s(time, bs = "cr", k = 20, by = condition) +
        (1 | participant),
    data = summary_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 8,
    cores = 8,
    file = "models/meta_gam.rds"
    )

# fitting the BGAMM
# meta_gamm <- brm(
#     # using by-participant SD of ERPs across trials
#     eeg_mean | se(eeg_sd) ~ condition +
#         s(participant, bs = "re") +
#         s(time, participant, bs = "fs", by = condition, m = 1),
#     data = summary_df,
#     family = gaussian(),
#     warmup = 2000,
#     iter = 5000,
#     chains = 8,
#     cores = 8,
#     control = list(adapt_delta = 0.95),
#     file = "models/meta_gamm.rds"
#     )
```

@@

We then compute the posterior probability of the slope for `condition` being above $0$ (@fig-post-prob-ratio, left). From this quantity, we then compute the ratio of posterior probabilities (i.e., $p/(1-p)$) and visualise the timecourse of this ratio superimposed with the conventional thresholds on evidence ratios (@fig-post-prob-ratio, right). Note that a ratio of 10 means that the probability of the difference being above 0 is 10 times higher than the probability of the difference not being above 0, given the data, the priors, and other model's assumptions. Thresholding the posterior probability ratio thus provides a model-based approach for estimating the onset and offset of M/EEG effects. An important advantage is that the proposed approach can be extended to virtually any model structure. Moreover, the approach provides both group-level and individual-level onset and offset estimates of M/EEG effects.

@@

```{r fig-bgamm-varying-effects, eval = FALSE, echo = FALSE, fig.width = 8, fig.asp = 0.5, fig.cap = "Posterior estimate of the EEG activity in each condition (left) and posterior estimate of the difference in EEG activity (right) according to the BGAMM."}
####################################################################
# TO-DO: plotting the onset/offset estimate per participant?
#################################################################

model <- meta_gamm
plot_post_preds(model = meta_gamm) + plot_post_slope(model = meta_gamm)

plot_post_test_ratio_individual <- function (model, sesoi = 0, threshold = 10) {

    # retrieving the number of posterior samples
    n_posterior_samples <- ndraws(model)

    # retrieving posterior draws and compute the difference per participant & time
    prob_y_above_0_indiv <- model$data %>%
        add_epred_draws(object = model) %>%
        data.frame() %>%
        dplyr::select(participant, time, condition, .epred, .draw) %>%
        pivot_wider(names_from = condition, values_from = .epred) %>%
        mutate(epred_diff = cond2 - cond1) %>%
        group_by(participant, time) %>%
        summarise(m = mean(epred_diff > 0 + sesoi), .groups = "drop") %>%
        mutate(
            prob_ratio = m / (1 - m),
            prob_ratio = ifelse(is.infinite(prob_ratio), n_posterior_samples, prob_ratio),
            prob_ratio = ifelse(prob_ratio == 0, 1 / n_posterior_samples, prob_ratio),
            above_thres = ifelse(prob_ratio >= threshold, 1, NA)
            )

    # plotting posterior probabilities per participant
    post_plot_indiv <- prob_y_above_0_indiv %>%
        ggplot(aes(x = time, y = m, group = participant, color = participant) ) +
        geom_hline(yintercept = 0.5, linetype = 2) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_vline(xintercept = true_offset, linetype = 2) +
        geom_line(alpha = 0.5) +
        labs(
            x = "Time (s)",
            y = expression(Pr(beta>0~"|"~data, model) ),
            title = "Posterior probabilities per participant"
            ) +
        ylim(c(0, 1) ) +
        theme(legend.position = "none")

    # plotting posterior probability ratio per participant (log scale)
    ratio_plot_indiv <- prob_y_above_0_indiv %>%
        ggplot(aes(x = time, y = prob_ratio, group = participant, color = participant) ) +
        geom_hline(yintercept = 1, linetype = "dashed") +
        geom_hline(yintercept = 10, linetype = "dashed", color = "darkred") +
        geom_hline(yintercept = 1 / 10, linetype = "dashed", color = "darkred") +
        geom_hline(yintercept = 100, linetype = "dashed", color = "orangered") +
        geom_hline(yintercept = 1 / 100, linetype = "dashed", color = "orangered") +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_vline(xintercept = true_offset, linetype = 2) +
        geom_line(alpha = 0.5) +
        geom_point(
          data = . %>% dplyr::filter(!is.na(above_thres) ),
          aes(y = threshold),
          colour = "darkgreen",
          shape = 16,
          size = 1.5,
          na.rm = TRUE
        ) +
        scale_y_log10(
          labels = label_log(digits = 2),
          limits = c(1 / n_posterior_samples, n_posterior_samples),
          breaks = c(1/1e4, 1/1e3, 1/1e2, 1/10, 1, 10, 1e2, 1e3, 1e4)
        ) +
        labs(
          x = "Time (s)",
          y = expression(Pr(beta>0~"|"~data) / (1 - Pr(beta>0~"|"~data) ) ),
          title = "Posterior probability ratios per participant"
        ) +
        theme(legend.position = "none")

    # combining plots
    post_plot_indiv + ratio_plot_indiv

}
```

@@

Note: Given the focus on the precision of onset/offset estimates, is it correct/relevant to mention that we are interested here in quantifying type-S and type-M (rather than type-1 or type-2) errors?

@@

We then compared the ability of the BGAMM to accurately estimate the onset and offset of the ERP difference to other widely-used methods. First, we conducted mass-univariate t-tests (thus treating each timestep independently) and identified the onset and offset of the ERP difference as the first and last values crossing an arbitrary significance threshold ($\alpha = 0.05$). We then followed the same approach but after applying different forms of multiplicity correction to the $p$-values. We compared two methods that control the FDR [i.e., `BH95`, @benjamini1995; and `BY01`, @benjamini2001], one method that controls the FWER [i.e., Holm–Bonferroni method, @holm1979], and two cluster-based permutation methods [permutation with a single cluster-forming threshold and threshold-free cluster enhancement, `TFCE`, @smith2009]. The `BH95`, `BY01`, and `Holm` corrections were applied to the p-values using the `p.adjust()` function in `R`. The cluster-based inference was implemented using a cluster-sum statistic of squared $t$-values, as implemented in `MNE-Python` [@gramfort2013], called via `r pkrt("reticulate")`. We also compared these estimates to the onset and offset as estimated using the binary segmentation algorithm, as implemented in `r pkrt("changepoint")`, and applied directly to the squared $t$-values [as in @rousselet_using_2025]. @fig-corrections illustrates the onsets and offsets estimated by each method on a single simulated dataset and shows that all methods systematically overestimate the true onset and underestimate the true offset.

@@

To assess the accuracy of group-level onset estimation, the various methods were compared using the bias (i.e., median(estimated-true)), median absolute error (MAE), root mean square error (RMSE), variance, and median absolute deviation (MAD) of onset/offset estimates computed on 1,000 simulated datasets. As in @rousselet_using_2025, each participant was assigned a random onset between 150 and 170ms. Whereas the present article focuses on one-dimensional signas (e.g., one M/EEG channel), we provice a 2D application in @sec-2D.

@@

To complement the simulation study, we evaluated the performance of the various methods on actual MEG data [decoding results from @nalborczyk:inprep]. In this study, we conducted time-resolved multivariate pattern analysis (MVPA, also known as decoding) of MEG data during reading tasks. As a result, we obtain a timecourse of decoding performance (ROC AUC), bounded between 0 and 1, for each participant ($N=32$). Next, we wanted to *test* whether the group-level average decoding accuracy is above chance (i.e., 0.5) at each timestep (@fig-decoding-data). To achieve this, we fitted a BGAM as introduced previously, but we replaced the $\mathrm{Normal}$ likelihood function by a $\mathrm{Beta}$ one to account for the bounded nature of AUC values (between 0 and 1) [for a tutorial on Beta regression, see @coretta2025].

Note that although we chose a basis dimension of $k=50$, which seems appropriate for the present data, this choice should be adapted according to the properties of the modelled data (e.g., signal-to-noise ratio, prior low-pass filtering, sampling rate) and should be assessed by the usual model checking tools (e.g., posterior predictive checks, see also @sec-basis). To better distinguish signal from noise, we also defined a region of practical equivalence [ROPE, @kruschke2017], defined as the chance level plus the standard deviation of the (group-level average) decoding performance during the baseline period.

@@

```{r tbl-simulation-results, eval = TRUE, echo = FALSE, results = "asis"}
#| tbl-cap: "Summary statistics of the onset and offset estimates for each method (ordered by the absolute value of the bias)."

@@

```{r tbl-reliability-results, eval = TRUE, echo = FALSE, results = "asis"}
#| tbl-cap: "Summary statistics of the onset and offset estimates for each method (ordered by the absolute value of the bias)."

# retrieving the reliability results
reliability_results <- readRDS(file = "results/reliability_results_100splits.rds")

# extracting the full data estimates
reliability_errors <- reliability_results %>%
    data.frame() %>%
    # keeping only the first and last timestep
    group_by(split_id, method, onset_offset) %>%
    summarise(
        value = if (onset_offset[1] == "onset") min(value) else max(value),
        .groups = "drop"
        ) %>%
    ungroup() %>%
    select(-split_id) %>%
    left_join(
        full_reliability_results,
        by = c("method", "onset_offset"),
        suffix = c("_split", "_full")
        ) %>%
    group_by(method, onset_offset) %>%
    mutate(error = value_split - value_full) %>%
    ungroup()

# summarising the results
reliability_summary <- reliability_errors %>%
    # mutate(
    #     error = case_when(
    #         grepl("onset", onset_offset) ~ value - true_onset,
    #         grepl("offset", onset_offset) ~ value - true_offset
    #         )
    #     ) %>%
    group_by(method, onset_offset) %>%
    summarise(
        median_error = median(error, na.rm = TRUE),
        median_abs_error = median(abs(error), na.rm = TRUE),
        rmse = sqrt(mean(error)^2),
        variance = var(error, na.rm = TRUE),
        mad = mad(error, na.rm = TRUE),
        .groups = "drop"
        ) %>%
    ungroup() %>%
    mutate(
        method = factor(
            x = method,
            levels = c(
                "raw_p", "pval_bh", "pval_by", "pval_holm",
                "cluster_mass", "cluster_tfce", "cpt", "brms"
                ),
            labels = c(
                "Raw p-value", "FDR BH95", "FDR BY01", "Holm", "Cluster mass",
                "TFCE", "Change point", "BGAM (new method)"
                ),
            )
        ) %>%
    mutate(onset_offset = factor(x = onset_offset, levels = c("onset", "offset") ) ) %>%
    data.frame()
